#Week2 excersize: Regression models  

Read and study data

The data is in .txt -format and is tab-separated.

```{r} 
setwd("~/Tyojuttuja/opinnot/GitHub/IODS-project")
data <- read.table(file = "~/Tyojuttuja/opinnot/GitHub/IODS-project/data/learning2014.txt", sep="\t")
#I changed the working directory back to project
```

The data table was studied with relevant r-functions such as 
str(), head() and dim().

The data is about learning attitudes and points got from an exam. In this dataset, there are 166 rows and 7 columns:
gender, age, attitude, deep, stra, surf, points.
Attitude is the learning attitude (1-5) and deep, stra and surf are means of different kind of learning-related questions (deep, surface or strategic, scales 1-5). Points-column is the points got from the exam.

Please find more detail about the data in here
http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-meta.txt

Summaries of all variables was studied with summary function:
Gender:
```{r}
summary(data$gender)
```
Age:
```{r}
summary(data$age)
```
Attitude:
```{r}
summary(data$attitude)
```
Deep questions:
```{r}
summary(data$deep)
```
Strategic questions:
```{r}
summary(data$stra)
```
Surface questions:
```{r}
summary(data$surf)
```
Exam points:
```{r}
summary(data$points)
```

```{r}
#Access the GGally and ggplot2 libraries
library(GGally)
library(ggplot2)

# Variable dispersions and correlations can be studied with ggpairs-function

data_correlations <- ggpairs(data, mapping =  aes(col= gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

data_correlations
```

From the data distributions, we can notice that some variables are about normally distributed (deep, stra and surf questions mainly). On the other hand age is right-tailed: most students are in their twenties but then there are a couple of 30-50 year old students. It looks like the exam points correlates mostly with attitude (r=0.43).This correlation is similar in males and females. The attitude does not correlate strongly with age (r<0.1)

```{r}
model <- lm(points ~ attitude + stra + surf, data = data)
summary(model)
```

Exam points correlated stronly with attitude (Beta=3.4 SE 0.6, P = 2*10-8) in a linear regression model with strategic (stra) and surface (surf) questions. This means that one unit increase in (mean) attitude (scale 1-5) will in increase your exam success by 3.4. The question variables did not significantly affect the exam points (P-values > 0.1).

```{r}
model2 <- lm(points ~ attitude, data = data)
summary(model2)
```
When the learning-variables were removed from the statistical model, the effect of attitude on exam points remained quite unchanged (Beta=3.5, SE 0.6, P = 4.2*10-9) meaning that one unit increase in (mean) attitude (scale 1-5) will in increase your exam success by 3.5. Attitude explaines 18.6% of the variability in the exam points (adjusted R-squared)